---
title: "MS Genotyping pipeline"
author: "Luciano Kalabric"
creation date: "2022-09-24"
last update: "2022-09-27"
output: html_document
---

```{r header, include=FALSE}
knitr::opts_chunk$set(echo = TRUE) # This a global statment
# Note:
# ===============================================================
# `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the outputs and plots

# General advises:
# ===============================================================
# 1) use "_" instead of "." in variable and function names;
# 2) separate each block of code with a comment and "=====" whenever possible
# 3) this list will continue growing as soon as I learn more about programming in R...
```

## MS Genotyping Pipeline

This is an effort to parse microsatellite trace files in fsa format together with epidemiological data to create useful data to be uploaded and analyzed at Online Program SpadeR https://chao.shinyapps.io/SpadeR/.

When you click the **Knit** button a document will be generated that includes both content as well as the output of any embedded R code chunks within the document. You can embed an R code chunk like this:

## Setup

Required packages installation/loading alongside other source codes

```{r setup}
# This piece of code turned out not to be useful since RStudio recognizes missing packages and asks for installation
# ===============================================================
list.of.packages <- c("Fragman","pacman")
new.packages <- list.of.packages[!(list.of.packages %in% installed.packages()[,"Package"])]
if(length(new.packages)) install.packages(new.packages)

# Loads package Fragman 
# ===============================================================
library('Fragman')

# Loads other source codes 
# ===============================================================
source("~/GitHub/peakscan/ladders_info.R") 		# ladders sizes
source("~/GitHub/peakscan/markers_info.R") 		# microsatellites info & sets (or batches)
source("~/GitHub/peakscan/research_data.R")		# epi & lab data

# General variables
# ===============================================================
threshold=100
my_sets <- unique(my_marker_sets$set)

# Output data to Streamlit
write.csv(my_sets, "data/my_sets.csv", row.names = FALSE)
write.csv(my_marker_sets, "data/my_marker_sets.csv", row.names = FALSE)
```
## Use Package Fragman

The core of the package Fragman and the workflow of the fragment analysis rely in 4 functions:
1) storing.inds (function in charge of reading the FSA or txt(CQS) files and storing them with a list structure)

```{r sorting_inds}
# It is a good practice to store all fsa files in a different folder. # Read just one folder at a time
#fsa_dir <- "~/GitHub/peakscan/fsa_set1"
#fsa_data <- storing.inds(fsa_dir, channels=5, lets.pullup=FALSE)
# Maybe we should think about saving each set of files in a different folder (i.e. fsa_set1_dir, fsa_set2_dir, etc.)
# ===============================================================
sets_genotyped <- c()
fsa_alldata <- list()
for (i in my_sets) {
  fsa_dir <- paste0("~/GitHub/peakscan/fsa_",i)
  if (length(list.files(fsa_dir, all.files = TRUE, include.dirs = TRUE, no.. = TRUE)) > 0) {
    list.files(fsa_dir, all.files = TRUE, include.dirs = TRUE, no.. = TRUE)
    # Reads the fsa files and stores them within a list structure
    # ===============================================================
    print(i)
    fsa_stored <- storing.inds(fsa_dir, channels=5, lets.pullup=FALSE)
    # Include here all step of Fragman instead of accumulate to analyze later!
    sets_genotyped <- c(sets_genotyped, i)
    fsa_alldata <- c(fsa_alldata, list(fsa_stored))
   } else { 
    print(paste("Directory",fsa_dir,"is empty!"))
  }
}
names(fsa_alldata) <- sets_genotyped
```

2) ladder.info.attach (uses the information read from the FSA files and a vector containing the ladder information (DNA size of the fragments) and matches the peaks from the channel where the ladder was run with the DNA sizes for all samples. Then loads such information in the R environment for the use of posterior functions

```{r ladder_info_attach}
# Create the list "list.data.covarrubias" indicating the ladder peaks position, height, weigth, correlation and error identified in each file
# ===============================================================
#ladder.info.attach(stored=fsa_data, ladder=gs600liz, ladd.init.thresh=threshold, prog=FALSE, draw=TRUE)
list_data_allcovarrubias <- list()
for (i in sets_genotyped) {
  print(paste("Matching ladder from samples of",i,"..."))
  ladder.info.attach(stored=fsa_alldata[[i]], ladder=gs600liz, ladd.init.thresh=threshold, prog=FALSE, draw=TRUE)
  list_data_allcovarrubias <- c(list_data_allcovarrubias, list(list.data.covarrubias))
}
names(list_data_allcovarrubias) <- sets_genotyped
print("Done.")
### Matching your ladder is a critical step and should only happen once per batch of samples read
```
OPTIONAL:
### If the ladder.info attach function detects some bad samples
### that you can correct them manually using
### the ladder.corrector() function
Here we identified all samples that we did not find a good ladder even though we tried the most liberal parameters:

```{r bad_ladder}
corro <- list()
bad <- c()
bad_samples <- c()
for (i in sets_genotyped) {
  # Extracts the correlation coeficiente between the ladder peak positions and sizes from the "list.data.covarrubia"
  print(paste("Testing ladder peaks from samples of",i,"..."))
  corro[[i]] <- unlist(lapply(list_data_allcovarrubias[[i]],function(x){x$corr}))
  # Filter files with bad ladder
  bad[[i]] <- which(corro[[i]] < .9999)
  # Extract the file names from the list bad to the vector bad_data and print the results, if any
  bad_samples[[i]] <- names(bad[[i]])
  if (length(bad_samples[[i]]) == 0) {
    print("All samples have a good ladder!")
  } else {
    print(paste("Here are the bad samples:",bad_samples[[i]]))
  }
  print("Done.")
}
```

You can continue your analysis without worrying about those samples or removing them. Alternatively, you may correct them manually by running the chuck below.

```{r ladder_corrector}
#ladder.corrector(fsa_data, bad_data, gs600liz, thresh=threshold)
```
3) overview & overview2 create friendly plots for any
number of individuals specified and can be used to design panels (overview2) for posterior automatic scoring (like licensed software does), or make manual scoring (overview)

```{r overview2}
# General loop per batch and marker
# =========================================================
for (i in sets_genotyped) {
  #markers_set_name = paste0("markers_",i)
  #assign(markers_info, subset(my_marker_sets, set == i))
  print(paste("Creating panel from markers of",i,"..."))
  markers_info <- subset(my_marker_sets, set == i)
  for (j in markers_info$marker) {
    print(paste("Working on marker",j,"..."))
    # Extract values from marker_info data.frame by marker
    ch = markers_info[markers_info$marker == j,]$ch
    x_min = markers_info[markers_info$marker == j,]$x_min
    x_max = markers_info[markers_info$marker == j,]$x_max
    my_panel <- overview2(my.inds=fsa_alldata[[i]], channel=ch, ladder=gs600liz, init.thresh=1000, xlim=c(x_min,x_max))
    marker_name = paste0("marker_",i,"_",j)
    assign(marker_name, my_painel)
    print("Done.")
    readline(prompt = "Press [enter] to continue.")
  }
}

readline(prompt = "Press [enter] to continue.")



# Design panels for smms set 1
# =========================================================
smms2_panel <- overview2(my.inds=fsa_data, channel=1, ladder=gs600liz, init.thresh=7000, xlim=c(219,295))
smms2_panel <- locator(type="p", pch=20, col="blue")$x
a <- score.markers (my.inds=fsa_data, channel=1, panel=smms2_panel, ladder=gs600liz, electro=FALSE)

smms13_panel <- overview2(my.inds=fsa_data, channel=4, ladder=gs600liz, init.thresh=7000, xlim=c(171,228))
smms13_panel <- locator(type="p", pch=20, col="red")$x
a <- score.markers (my.inds=fsa_data, channel=4, panel=smms13_panel, ladder=gs600liz, electro=FALSE)

smms16_panel <- overview2(my.inds=fsa_data, channel=2, ladder=gs600liz, init.thresh=7000, xlim=c(201,258))
smms16_panel <- locator(type="p", pch=20, col="green")$x
a <- score.markers (my.inds=fsa_data, channel=2, panel=smms16_panel, ladder=gs600liz, electro=FALSE)

```

